---
title: "Visual - Analytics 1st Deliverable"
output:
  html_document:
    df_print: paged
---



#1. We import the LSMS-ISA Nigeria Wave 3 Dataset from my personal repository
```{r getData, eval=TRUE}
library(rio)
rm(list = ls())
nigeria_w3<-read.csv('https://github.com/Sebastian-Wood-Visual-Analytics/deliverable1/raw/main/NGA_W3_EPAR.csv')
View(nigeria_w3)
```

#2. #Select some of the variables and create a new Dataset
```{r}

library(tidyverse)
nigeria_w3 <- select(nigeria_w3, hhid, zone, rural, daily_percap_cons, months_food_insec)
```

#3. Data Types of each variable:
```{r}
names(nigeria_w3)
```

#4.Data types of each variable
```{r}
#this 'width = 70,strict.width='cut' means
# you do not want to see more than 70 characters per row.

str(nigeria_w3,width = 70,strict.width='cut')
#You see different data types in the result: num (numbers in R when decimal values are detected). The ones that say chr are strings. You see several variables identified as factor or ordered factor, which are categorical variables.
```


We start constructing the Univariate Plot months_food_insec
```{r}
#You can not get a clear idea of what a data table has, so a simple frequency table is the first tool to see what these nominal data are telling us:
# absolute values
absoluteT=table(nigeria_w3$months_food_insec,
                exclude = 'nothing') #include all values!
absoluteT

#The absolute count can be transformed into percent or relative frequencies:
# relative values
prop.table(absoluteT)

#We should try using all the information possible, then it is a good idea to label the missing values:
names(absoluteT)[8]='Unknown'

#If you prefer that the range of the proportions be translated into percents you can do:
propT=prop.table(absoluteT)*100
#you get:
propT

# the pie plots the table:
pie(absoluteT)

#Let’s turn our pie into a bar chart using: ggplot2.
```


```{r}
#The basic steps to prepare a visual in ggplot2 are: 
#Turn the frequency table into a data frame:
# as data frame
(tableFreq=as.data.frame(absoluteT))

#The previous frequency table can be improved:
# renaming data frame columns
names(tableFreq)=c("monthfoodinsecure","Count")
# adding percents:
tableFreq$Percent=as.vector(propT)
# then, you have:
tableFreq

#This data frame has the three columns, we will use second for our bars.
# b. Call the library (you may need to install it first):
library(ggplot2)

#c. Create the base object, which is not a plot, just informing the main variables:
#base GGPLOT2 starts with a "base", telling WHAT VARIABLES TO PLOT
base= ggplot(data = tableFreq, 
             aes(x = monthfoodinsecure,
                 y = Percent)) 

# D. On top of the previous object, add the layer that produces the main plots (the next layers will add or customize elements in the plot):
plot1 = base + geom_bar(fill ="gray",
                        stat = 'identity') 
plot1
```



```{r}
# You have a default bar plot now. From here, you start adding more layers that will change the default.

#Add titles:
titleText='Nigeria 2015-16: Food Insecurity for Agricultural Small-Scale Producers'
sourceText='Source: Evans School Policy Analysis and Research Group (EPAR).'
x_axis='Number of Months Food Insecure'
y_axis='Percentage of Total Households surveyed (%)'

plot2 = plot1 + labs(title=titleText,
                     x =x_axis,
                     y =y_axis,
                     caption = sourceText) 
plot2

#Add the reference lines: Let’s use the percent column and annotate to show a particular value:
base= ggplot(data = tableFreq, 
             aes(x = monthfoodinsecure,
                 y = Percent)) 

plot1 = base + geom_bar(fill ="gray",
                        stat = 'identity') 

plot2 = plot1 + labs(title=titleText,
                     x =x_axis, 
                     y = y_axis,
                     caption = sourceText)

plot3 = plot2 
#+ geom_hline(yintercept = 12.5, #where
 #                          linetype="dashed", 
  #                         size=1.5, #thickness
   #                        alpha=0.5) #transparency
plot3
```


```{r}
library(scales) # for "unit_format""

# customize Y axis
plot4 = plot3 + scale_y_continuous(breaks=c(0,25,50,75,100),
                                   limits = c(0,80), 
                                   labels=unit_format(suffix = '%')) 
plot4
```

* Title/subtitle positions:
```{r}
#positions: 0 left / 1 right / 0.5 center
plot5 = plot4 + theme(plot.caption = element_text(hjust = 0), 
                      plot.title = element_text(hjust = 0.5))
plot5
```


```{r}
paste0(round(tableFreq$Percent,2), '%')
# creating labels:
LABELS=paste0(round(tableFreq$Percent,2), '%')
###
plot6 = plot5 + geom_text(vjust=0, #hjust if flipping
                          size = 3,
                          aes(y = Percent ,
                              label = LABELS))
plot6# + coord_flip() # wanna flip the plot?

```




You may also need to reduce more the data-ink ratio, so I recommend changing the theme to **classic**. This and the object _localesOrd_ can be added to the previous base:
```{r}
# redo:
LABELS=paste0(round(tableFreq$Percent,2), '%')
base= ggplot(data = tableFreq, 
             aes(x = monthfoodinsecure,y = Percent)) 
## adding these to base
#base= base + scale_x_discrete(limits=localesOrd) 
base= base + theme_classic()
##
plot1 = base + geom_bar(fill ="gray",
                        stat = 'identity') 
plot2 = plot1 + labs(title=titleText,
                     x =x_axis, 
                     y = y_axis,
                     caption = sourceText)
plot3 = plot2 # + geom_hline(yintercept = 12.5, 
              #             linetype="dashed", 
               #            size=1.5, 
                #           alpha=0.5)
plot4 = plot3 + scale_y_continuous(breaks=c(0,25,50,75),
                                 limits = c(0, 80),
                                 # expand = c(0, 0), #try?
                            labels=unit_format(suffix = '%')) 

plot5 = plot4 + theme(plot.caption = element_text(hjust = 0),
                      plot.title = element_text(hjust = 0.5))


plot6 = plot5 + geom_text(vjust=0, #hjust if flipping
                          size = 4,#fontface = "bold",
                          aes(y = Percent,
                              label = LABELS))
plot6 #+ coord_flip() # wanna flip the plot?
```

