---
title: "Visual - Analytics 1st Deliverable"
output:
  html_document:
    df_print: paged
---



#1. We import the LSMS-ISA Nigeria Wave 3 Dataset from my personal repository
```{r getData, eval=TRUE}
library(rio)
rm(list = ls())
nigeria_w3<-read.csv('https://github.com/Sebastian-Wood-Visual-Analytics/deliverable1/raw/main/NGA_W3_EPAR.csv')
View(nigeria_w3)
```

#2. #Select some of the variables and create a new Dataset
```{r}

library(tidyverse)
nigeria_w3 <- select(nigeria_w3, hhid, zone, rural, daily_percap_cons, months_food_insec)
```

#3. Data Types of each variable:
```{r}
names(nigeria_w3)
```

#4.Data types of each variable
```{r}
#this 'width = 70,strict.width='cut' means
# you do not want to see more than 70 characters per row.

str(nigeria_w3,width = 70,strict.width='cut')
#You see different data types in the result: num (numbers in R when decimal values are detected). The ones that say chr are strings. You see several variables identified as factor or ordered factor, which are categorical variables.
```


We start constructing the Univariate Plot months_food_insec
```{r}
#You can not get a clear idea of what a data table has, so a simple frequency table is the first tool to see what these nominal data are telling us:
# absolute values
absoluteT=table(nigeria_w3$months_food_insec,
                exclude = 'nothing') #include all values!
absoluteT

#The absolute count can be transformed into percent or relative frequencies:
# relative values
prop.table(absoluteT)

#We should try using all the information possible, then it is a good idea to label the missing values:
names(absoluteT)[8]='Unknown'

#If you prefer that the range of the proportions be translated into percents you can do:
propT=prop.table(absoluteT)*100
#you get:
propT

# the pie plots the table:
pie(absoluteT)

#Let’s turn our pie into a bar chart using: ggplot2.
```


```{r}
#The basic steps to prepare a visual in ggplot2 are: 
#Turn the frequency table into a data frame:
# as data frame
(tableFreq=as.data.frame(absoluteT))

#The previous frequency table can be improved:
# renaming data frame columns
names(tableFreq)=c("monthfoodinsecure","Count")
# adding percents:
tableFreq$Percent=as.vector(propT)
# then, you have:
tableFreq

#This data frame has the three columns, we will use second for our bars.
# b. Call the library (you may need to install it first):
library(ggplot2)

#c. Create the base object, which is not a plot, just informing the main variables:
#base GGPLOT2 starts with a "base", telling WHAT VARIABLES TO PLOT
base= ggplot(data = tableFreq, 
             aes(x = monthfoodinsecure,
                 y = Percent)) 

# D. On top of the previous object, add the layer that produces the main plots (the next layers will add or customize elements in the plot):
plot1 = base + geom_bar(fill ="gray",
                        stat = 'identity') 
plot1
```



```{r}
# You have a default bar plot now. From here, you start adding more layers that will change the default.

#Add titles:
titleText='Nigeria 2015-16: Food Insecurity for Agricultural Small-Scale Producers'
sourceText='Source: Evans School Policy Analysis and Research Group (EPAR).'
x_axis='Number of Months Food Insecure'
y_axis='Percentage of Total Households surveyed (%)'

plot2 = plot1 + labs(title=titleText,
                     x =x_axis,
                     y =y_axis,
                     caption = sourceText) 
plot2

#Add the reference lines: Let’s use the percent column and annotate to show a particular value:
base= ggplot(data = tableFreq, 
             aes(x = monthfoodinsecure,
                 y = Percent)) 

plot1 = base + geom_bar(fill ="gray",
                        stat = 'identity') 

plot2 = plot1 + labs(title=titleText,
                     x =x_axis, 
                     y = y_axis,
                     caption = sourceText)

plot3 = plot2 + geom_hline(yintercept = 12.5, #where
                           linetype="dashed", 
                           size=1.5, #thickness
                           alpha=0.5) #transparency
plot3
```


```{r}
library(scales) # for "unit_format""

# customize Y axis
plot4 = plot3 + scale_y_continuous(breaks=c(0,25,50,75,100),
                                   limits = c(0,80), 
                                   labels=unit_format(suffix = '%')) 
plot4
```

* Title/subtitle positions:
```{r}
#positions: 0 left / 1 right / 0.5 center
plot5 = plot4 + theme(plot.caption = element_text(hjust = 0), 
                      plot.title = element_text(hjust = 0.5))
plot5
```


```{r}
paste0(round(tableFreq$Percent,2), '%')
# creating labels:
LABELS=paste0(round(tableFreq$Percent,2), '%')
###
plot6 = plot5 + geom_text(vjust=0, #hjust if flipping
                          size = 3,
                          aes(y = Percent ,
                              label = LABELS))
plot6# + coord_flip() # wanna flip the plot?

```




You may also need to reduce more the data-ink ratio, so I recommend changing the theme to **classic**. This and the object _localesOrd_ can be added to the previous base:
```{r}
# redo:
LABELS=paste0(round(tableFreq$Percent,2), '%')
base= ggplot(data = tableFreq, 
             aes(x = monthfoodinsecure,y = Percent)) 
## adding these to base
#base= base + scale_x_discrete(limits=localesOrd) 
base= base + theme_classic()
##
plot1 = base + geom_bar(fill ="gray",
                        stat = 'identity') 
plot2 = plot1 + labs(title=titleText,
                     x =x_axis, 
                     y = y_axis,
                     caption = sourceText)
plot3 = plot2 + geom_hline(yintercept = 12.5, 
                           linetype="dashed", 
                           size=1.5, 
                           alpha=0.5)
plot4 = plot3 + scale_y_continuous(breaks=c(0,25,50,75),
                                 limits = c(0, 80),
                                 # expand = c(0, 0), #try?
                            labels=unit_format(suffix = '%')) 

plot5 = plot4 + theme(plot.caption = element_text(hjust = 0),
                      plot.title = element_text(hjust = 0.5))


plot6 = plot5 + geom_text(vjust=0, #hjust if flipping
                          size = 4,#fontface = "bold",
                          aes(y = Percent,
                              label = LABELS))
plot6 #+ coord_flip() # wanna flip the plot?
```



##Delivarable 2 Bivariate relationships : Categorical-Categorical
```{r}
#Preparing Contigency Tables
zones_monthsfinsecure=table(nigeria_w3$zone,nigeria_w3$months_food_insec,
                            exclude = '')
zones_monthsfinsecure


```

When a table tries to hypothesize a relationship, you should have the independent variable in the columns (food insecurity), and the dependent one (zones) in the rows. Interpretation is difficult when you have counts so it is better to have percents. Percents should be computed by column to see how the levels of the dependent variable varies by each level of the independent one (reading along rows):

```{r}
# computing column percent from contingency table
library(magrittr) # for %>%
(zones_monthsfinsecure.mgCol=prop.table(zones_monthsfinsecure,
                            margin = 2) #2 means by column
                             %>%round(.,3))
```
 As before, we need to turn this table into a data frame:
```{r}
#making a data frame from contingency table
(zones_monthsfinsecureDF=as.data.frame(zones_monthsfinsecure))

#We also have the table with marginal percents by column:

as.data.frame(zones_monthsfinsecure.mgCol)

#We should simply add the last column to the data frame of counts.

zones_monthsfinsecureDF$share=as.data.frame(zones_monthsfinsecure.mgCol)[,3]
zones_monthsfinsecureDF

#We can change the names of the previous data frame:

names(zones_monthsfinsecureDF)[1:3]=c("zones","m_foodins","counts")

#then
zones_monthsfinsecureDF
```

We will use ggplot to represent the contingency table:
```{r}

library(ggplot2)
base1=ggplot(data=zones_monthsfinsecureDF, 
             aes(x=m_foodins,
                 y=share,
                 fill=zones)) # fill brings a legend
```

```{r}
#Then, you play with some positions for the bar. First, the dodge style:

barDodge= base1 +  geom_bar(stat="identity",
                            position ='dodge') 
barDodge 
```

Te second is the stack style:
```{r}
barStacked = base1 + geom_bar(stat = "identity",
                              position = 'stack')#default
barStacked 

```

The stacked version will help more than the dodged one as it reveals better the values in the contingency table:
```{r}
zones_monthsfinsecure.mgCol
```

So, we continue with adding some elements to this one:
```{r}
library(scales)
#annotating
barStackedAnn= barStacked + geom_text(size = 3,# check below:
                             position = position_stack(vjust = 0.5),# center
                             aes(label=percent(share,accuracy = 0.1)))# percent format

barStackedAnn = barStackedAnn + scale_y_continuous(labels = percent)

barStackedAnn
```


Since the precinct is nominal, and you see some marked differences along the percents, you can reorder the precinct blocks with reorder():
```{r}
base1=ggplot(data=zones_monthsfinsecureDF, 
             aes(x=m_foodins, y=share,
                 fill=reorder(zones,share))) ## reordering

barStacked = base1 + geom_bar(stat = "identity",
                              position = 'stack')
barStacked= barStacked + geom_text(size = 3,
                             position = position_stack(vjust = 0.5),
                             aes(label=percent(share,accuracy = 0.1)))

barStacked = barStacked + scale_y_continuous(labels = percent)

barStacked
```

```{r}
#Add titles:
titleText='Nigeria 2015-16: Food Insecurity for Agricultural SSP by Geographic Zones'
sourceText='Source: Evans School Policy Analysis and Research Group (EPAR).'
x_axis='Number of Months Food Insecure'
y_axis='Percentage of Total Households surveyed (%)'
check="Zones"
barStacked = barStacked + labs(title=titleText,
                     x =x_axis,
                     y =y_axis,
                     caption = sourceText,
                     fill =check) 
barStacked
```

